---
title: "Appendix B (categorical surprisal)"
author: "S. Slaats"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load packages
library(emmeans)
library(plyr)
library(dplyr)
library(lme4)
library(gridExtra)
library(grid)
library(lmerTest)
library(car)
library(MAP)
library(broom)
library(tidyverse)
library(cAIC4)
library(writexl)
library(ggplot2)
library(patchwork)

emm_options(lmerTest.limit = 5000)
```

# Model selection
## Loading the data
```{r loading, include=TRUE}

# working directory
setwd("C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/")

# load data 
data <- read.csv('3 - SPR main/analysis/readingtimes_88pp_residuals_manual-outliers-REVIEWS.csv', header = TRUE)
stiminfo <- read.csv('5 - metadata/frequency_length_position.csv', header = TRUE)

# set factors
data$agreement <- factor(data$agreement, levels=c('correct','incorrect'))
data$surprisal <- factor(data$surprisal, levels=c('low','high'))
data$correct_number <- factor(data$correct_number, levels=c("singular", "plural"))
data$set_number <- factor(data$set_number)

fmla <- 'logRTresidual ~ agreement * surprisal + agreement * correct_number + 
                              (1 + agreement * surprisal + agreement * correct_number | UserId) + 
                              (1 + agreement * surprisal | set_number )'

fixed <-  'logRTresidual ~ agreement * surprisal + agreement * correct_number'

fixed_effects <- list('logRTresidual ~ agreement * surprisal + agreement * correct_number', # 1
                      'logRTresidual ~ surprisal + agreement * correct_number',             # 2
                      'logRTresidual ~ agreement * correct_number',                         # 3
                      'logRTresidual ~ agreement * surprisal + correct_number',             # 4
                      'logRTresidual ~ agreement * surprisal',                              # 5
                      'logRTresidual ~ surprisal + agreement + correct_number',             # 6
                      'logRTresidual ~ agreement + correct_number',                         # 7
                      'logRTresidual ~ surprisal + correct_number',                         # 8
                      'logRTresidual ~ surprisal + agreement',                              # 9
                      'logRTresidual ~ surprisal',                                          # 10
                      'logRTresidual ~ agreement',                                          # 11
                      'logRTresidual ~ correct_number',                                     # 12
                      'logRTresidual ~ ')                                                     # 13

participant_effects <- c('(1 + agreement * surprisal + agreement * correct_number | UserId)',
                         '(1 + agreement * surprisal + correct_number | UserId)',
                         '(1 + agreement * correct_number + surprisal | UserId)',
                         '(1 + agreement + correct_number + surprisal | UserId)',
                         '(1 + agreement * correct_number | UserId)',
                         '(1 + agreement + correct_number | UserId)',
                         '(1 + agreement * surprisal | UserId)',
                         '(1 + agreement + surprisal | UserId)',
                         '(1 + surprisal + correct_number | UserId)',
                         '(1 + agreement | UserId)',
                         '(1 + surprisal | UserId)',
                         '(1 + correct_number | UserId)',
                         '(1 | UserId)'
                         )

item_effects <- c('(1 + agreement * surprisal | set_number )',
                  '(1 + agreement + surprisal | set_number )',
                  '(1 + surprisal | set_number )',
                  '(1 + agreement | set_number )',
                  '(1 | set_number)'
                  )

# the random effects formulas
tmp_fmlas <- list()
for (randi in 1:length(item_effects)) {
  for (randp in 1:length(participant_effects)) {
    
    tmp_fmla <- sprintf('%s + %s + %s', fixed,  item_effects[[randi]], participant_effects[[randp]])
    tmp_fmlas <- c(tmp_fmlas, tmp_fmla)
    
  }
}
```
## Helper functions
```{r permutation settings, include=TRUE}
n_permutations <- 1000
seed <- 42
set.seed(seed)

# formula
permute_lmer <- function(model_orig,
                         data,
                         n_permutations,
                         group_var = "UserId",
                         response_var = "logRTresidual",
                         REML=FALSE,
                         verbose=TRUE){
  
  formula <- formula(model_orig)
  coef_names <- names(fixef(model_orig))
  obs_coefs <- fixef(model_orig)
  
  # storage matrix (rows = perms, cols = fixed effects)
  perm_matrix <- matrix(NA, nrow = n_permutations, ncol = length(coef_names))
  colnames(perm_matrix) <- coef_names
  
  for (i in 1:n_permutations) {
    # shuffle labels
    data_perm <- data %>%
      group_by(.data[[group_var]]) %>%
        mutate(
          !!response_var := sample(.data[[response_var]])
        ) %>%
        ungroup()
    # Fit model to permuted data
    tryCatch({
      perm_model <- lmer(formula, data = data_perm, control=lmerControl(optimizer="nloptwrap", optCtrl=list(maxfun=1e6)), REML = REML)
      perm_matrix[i, ] <- fixef(perm_model)
    }, error = function(e) {
      # Skip failed fits
    })
    
    if (verbose && i %% 50 == 0) {
      message(sprintf("Completed %d / %d permutations", i, n_permutations))
    }
    
  }
  
  perm_matrix <- perm_matrix[complete.cases(perm_matrix), , drop = FALSE]
  
  p_values <- sapply(seq_along(obs_coefs), function(j) {
    mean(abs(perm_matrix[, j]) >= abs(obs_coefs[j]))
  })
  names(p_values) <- coef_names
  
  return(list(
    observed_coefs = obs_coefs,
    permuted_coefs = perm_matrix,
    p_values = p_values
  ))

}

plot_permutation_densities <- function(result_list) {

  permuted_df <- as.data.frame(result_list$permuted_coefs)
  observed <- result_list$observed_coefs
  pvals <- result_list$p_values
  
  coef_names <- setdiff(names(permuted_df), "(Intercept)")

  plots <- lapply(coef_names, function(coef_name) {
    ggplot(permuted_df, aes(x = .data[[coef_name]])) +
      geom_density(fill = "skyblue", alpha = 0.5) +
      geom_vline(xintercept = observed[coef_name], color = "red", linewidth = 1) +
      labs(
        title = sprintf("%s\n(p = %.3f)", coef_name, pvals[coef_name]),
        x = "Coefficient value",
        y = "Density"
      ) +
      theme_minimal(base_size=10) +
      theme(plot.title = element_text(size = 10, hjust = 0.5))  # Center and shrink subplot titles
  })

    # Combine all plots into one figure
  combined_plot <- wrap_plots(plots, ncol = length(plots)) +
    plot_annotation(title = "Permutation Test Distributions for Fixed Effects",
      theme = theme(plot.title = element_text(hjust = 0.5, size = 14))
    )

  return(combined_plot)
}

```

```{r Function count number of random parameters, include=TRUE}
count_random <- function(mdl) {

  vc_list <- VarCorr(mdl)

  # Count variance components
  sum(sapply(vc_list, function(v) {
    n_std  <- length(attr(v, "stddev"))
    n_corr <- if (!is.null(attr(v, "correlation"))) {
                length(attr(v, "correlation"))
              } else 0L
    n_std + n_corr
  }))
}

```

```{r Function random effects, include=TRUE}
fit_random <- function(tmp_fmla, data) {
  model <- lmer(tmp_fmla, data, control=lmerControl(optimizer="nloptwrap", optCtrl=list(maxfun=2e5)), REML=FALSE)

}

fit_fixed <- function(tmp_fixed, random, data) {
  tmp_fmla <- eval(parse(text=paste(tmp_fixed, " + ", random)))
  model <- lmer(tmp_fmla, data, control=lmerControl(optimizer="nloptwrap", optCtrl=list(maxfun=1e6)), REML=FALSE)
  
}
```

## Random effects - estimation of all models
### Word index 5 (target)
```{r Random effects word index 5, include=TRUE, message=FALSE, warning=FALSE}
models5 <- lapply(tmp_fmlas, fit_random, subset(data, word_index==5))
saveRDS(models5, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models5CAT.Rdata')
```

### Word index 6 (spill-over 1)
```{r Random effects word index 6, include=TRUE, message=FALSE, warning=FALSE}
models6 <- lapply(tmp_fmlas, fit_random, subset(data, word_index==6))
saveRDS(models6, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models6CAT.Rdata')
```

### Word index 6 (spill-over 2)
```{r Random effects word index 7, include=TRUE, message=FALSE, warning=FALSE}
models7 <- lapply(tmp_fmlas, fit_random, subset(data, word_index==7))
saveRDS(models7, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models7CAT.Rdata')
```

### Word index 6 (spill-over 3)
```{r Random effects word index 8, include=TRUE, message=FALSE, warning=FALSE}
models8 <- lapply(tmp_fmlas, fit_random, subset(data, word_index==8))
saveRDS(models8, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models8CAT.Rdata')
```

## Evaluating the random models, fixed effects estimation & model selection
### Word index 5 (target)
```{r random models index 5, include=TRUE, message=FALSE, warning=FALSE}
models5 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models5CAT.Rdata')

good_models5 <- list()
for (mdli in 1:length(models5)) {
  mdl <- models5[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    good_models5 <- append(good_models5, mdl)
  }
}

# now we sort them for number of variables
# and their AIC -- we choose the model with the *highest number of variables* and *lowest AIC* combination
rand_counts <- vapply(good_models5, count_random, integer(1))
AIC_vals <- sapply(good_models5, AIC)
nvar = order(-rand_counts)
aics = order(AIC_vals, decreasing=FALSE)

# now we take the sum of the position of each model
var_ix = match(1:length(good_models5), nvar)
aic_ix = match(1:length(good_models5), aics)

score <- var_ix + aic_ix 

sorted_models5 <- good_models5[order(score)]

# we pick the largest one
start_model5 <- sorted_models5[[1]]
print(formula(start_model5))

# now we reduce the fixed effects - STEP keeps reducing random effects too
# so we go back to manual selection
fixed5 <- lapply(fixed_effects, fit_fixed, "(1 + agreement * surprisal | set_number) + (1 + agreement | 
    UserId)", subset(data, word_index==5))

# save the fixed models
saveRDS(fixed5, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/fixed5CAT.Rdata')

final_models5 <- list()
for (mdli in 1:length(fixed5)) {
  mdl <- fixed5[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    final_models5 <- append(final_models5, mdl)
  }
}

fixed_aics = sapply(final_models5, AIC)
best_model5 = final_models5[order(fixed_aics, decreasing=FALSE)][[1]]
print(formula(best_model5))
summary(best_model5)
saveRDS(best_model5, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model5CAT.Rdata')
```
#### Permutation test
```{r permutation5, include=TRUE, message=FALSE, warning=FALSE}
best_model5 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model5CAT.Rdata')
perm5 <- permute_lmer(best_model5,
                      subset(data, word_index==5),
                      n_permutations,
                      verbose=TRUE)
saveRDS(perm5, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/perm55CAT.Rdata')
```

```{r word index 5 viz permutation, fig.width=7.5,fig.height=4, include=TRUE, message=FALSE, warning=FALSE}
plots5 <- plot_permutation_densities(perm5)
print(plots5)
print(perm5$p_values)
```
#### Simple effects
```{r word index 5 EMMEANS, include=TRUE, message=FALSE, warning=FALSE}
# agreement by number
mdl5_agr_by_num <- emmeans(best_model5, specs='agreement', by='correct_number')

# number by agreement
mdl5_num_by_agr <- emmeans(best_model5, specs='correct_number', by='agreement')

# get results without adjustment
mdl5_agr_by_num_raw <- summary(pairs(mdl5_agr_by_num), adjust='none')
mdl5_num_by_agr_raw <- summary(pairs(mdl5_num_by_agr), adjust='none')

# extract pvalues
all_pvals <- c(mdl5_agr_by_num_raw$p.value, mdl5_num_by_agr_raw$p.value)
adj_pvals <- p.adjust(all_pvals, method='bonferroni')

# extract pvalues
mdl5_res_pair <- rbind(mdl5_agr_by_num_raw, mdl5_num_by_agr_raw)
mdl5_res_pair$adj.p.value <- adj_pvals
summary(mdl5_res_pair)
``` 

### Word index 6 (spill-over 1)
```{r Check random models index 6, include=TRUE, message=FALSE, warning=FALSE}
models6 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models6CAT.Rdata')

good_models6 <- list()
for (mdli in 1:length(models6)) {
  mdl <- models6[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    good_models6 <- append(good_models6, mdl)
  }
}

# now we sort them
rand_counts <- vapply(good_models6, count_random, integer(1))
AIC_vals <- sapply(good_models6, AIC)
nvar = order(-rand_counts)
aics = order(AIC_vals, decreasing=FALSE)

# now we take the sum of the position of each model
var_ix = match(1:length(good_models6), nvar)
aic_ix = match(1:length(good_models6), aics)
score <- var_ix + aic_ix 

sorted_models6 <- good_models6[order(score)]

# we pick the best/largest one
start_model6 <- sorted_models6[[1]]
print(formula(start_model6))

# now we reduce the fixed effects - STEP keeps reducing random effects too
# so we go back to manual selection
fixed6 <- lapply(fixed_effects, fit_fixed, "(1 + surprisal | set_number) + (1 + agreement | UserId)", subset(data, word_index==6))

# save the fixed models
saveRDS(fixed6, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/fixed6CAT.Rdata')

# now we check whether they converged
final_models6 <- list()
for (mdli in 1:length(fixed6)) {
  
  mdl <- fixed6[[mdli]]
  
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    final_models6 <- append(final_models6, mdl)
  }
}

# they did, so we pick the model with the highest AIC
fixed_aics = sapply(final_models6, AIC)
best_model6 = final_models6[order(fixed_aics, decreasing=FALSE)][[1]]
print(formula(best_model6))
summary(best_model6)
saveRDS(best_model6, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model6CAT.Rdata')
```
#### Permutation test
```{r permutation6, include=TRUE, message=FALSE, warning=FALSE}
# load the best model
best_model6 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model6CAT.Rdata')
perm6 <- permute_lmer(best_model6,
                      subset(data, word_index==6),
                      n_permutations,
                      verbose=TRUE)
saveRDS(perm6, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/perm6CAT.Rdata')
```

```{r word index 6 viz permutation, fig.width=7.5,fig.height=4, include=TRUE, message=FALSE, warning=FALSE}
plots6 <- plot_permutation_densities(perm6)
print(plots6)
print(perm6$p_values)
```

#### Simple effects
```{r word index 6 EMMEANS, include=TRUE, message=FALSE, warning=FALSE}
# agreement by number
mdl6_agr_by_num <- emmeans(best_model6, specs='agreement', by='correct_number')

# number by agreement
mdl6_num_by_agr <- emmeans(best_model6, specs='correct_number', by='agreement')


# get results without adjustment
mdl6_agr_by_num_raw <- summary(pairs(mdl6_agr_by_num), adjust='none')
mdl6_num_by_agr_raw <- summary(pairs(mdl6_num_by_agr), adjust='none')

# extract pvalues
all_pvals <- c(mdl6_agr_by_num_raw$p.value, mdl6_num_by_agr_raw$p.value)
adj_pvals <- p.adjust(all_pvals, method='bonferroni')

# extract pvalues
mdl6_res_pair <- rbind(mdl6_agr_by_num_raw, mdl6_num_by_agr_raw)
mdl6_res_pair$adj.p.value <- adj_pvals
summary(mdl6_res_pair)
``` 

### Word index 7 (spill-over 2)
```{r Check random models index 7, include=TRUE, message=FALSE, warning=FALSE}
models7 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models7CAT.Rdata')

good_models7 <- list()
for (mdli in 1:length(models7)) {
  mdl <- models7[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    good_models7 <- append(good_models7, mdl)
  }
  
}

# now we sort them
rand_counts <- vapply(good_models7, count_random, integer(1))
AIC_vals <- sapply(good_models7, AIC)
nvar = order(-rand_counts)
aics = order(AIC_vals, decreasing=FALSE)

# now we take the sum of the position of each model
var_ix = match(1:length(good_models7), nvar)
aic_ix = match(1:length(good_models7), aics)
score <- var_ix + aic_ix 

sorted_models7 <- good_models7[order(score)]

# we pick the best/largest one
start_model7 <- sorted_models7[[1]]
print(formula(start_model7))

# now we reduce the fixed effects - STEP keeps reducing random effects too
# so we go back to manual selection
fixed7 <- lapply(fixed_effects, fit_fixed, "(1 + agreement + surprisal | set_number) + (1 | UserId)", subset(data, word_index==7))

# save the fixed models
saveRDS(fixed7, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/fixed7CAT.Rdata')

# now we check whether they converged
final_models7 <- list()
for (mdli in 1:length(fixed7)) {
  mdl <- fixed7[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    final_models7 <- append(final_models7, mdl)
  }
}

# there was only one model left (code not necessary)
fixed_aics = sapply(final_models7, AIC)
best_model7 = final_models7[order(fixed_aics, decreasing=FALSE)][[1]]
print(formula(best_model7))
summary(best_model7)
saveRDS(best_model7, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model7CAT.Rdata')
```
#### Permutation test
```{r permutation7, include=TRUE, message=FALSE, warning=FALSE}
best_model7 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model7CAT.Rdata')
perm7 <- permute_lmer(best_model7,
                      subset(data, word_index==7),
                      n_permutations,
                      verbose=TRUE)
saveRDS(perm7, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/perm7CAT.Rdata')
```

```{r word index 7 viz permutation, fig.width=5,fig.height=4, include=TRUE, message=FALSE, warning=FALSE}
plots7 <- plot_permutation_densities(perm7)
print(plots7)
print(perm7$p_values)
```

### Word index 8 (spill-over 3)
```{r Check random models index 8, include=TRUE, message=FALSE, warning=FALSE}
models8 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/models8CAT.Rdata')

good_models8 <- list()
for (mdli in 1:length(models8)) {
  mdl <- models8[[mdli]]
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    good_models8 <- append(good_models8, mdl)
  }
}

# now we sort them
rand_counts <- vapply(good_models8, count_random, integer(1))
AIC_vals <- sapply(good_models8, AIC)
nvar = order(-rand_counts)
aics = order(AIC_vals, decreasing=FALSE)

# now we take the sum of the position of each model
var_ix = match(1:length(good_models8), nvar)
aic_ix = match(1:length(good_models8), aics)
score <- var_ix + aic_ix 

sorted_models8 <- good_models8[order(score)]

# we pick the best/largest one
start_model8 <- sorted_models8[[1]]
print(formula(start_model8))

# now we reduce the fixed effects - STEP keeps reducing random effects too
# so we go back to manual selection
fixed8 <- lapply(fixed_effects, fit_fixed, "(1 + agreement + surprisal | set_number) + (1 + agreement | 
    UserId)", subset(data, word_index==8))

# save the fixed models
saveRDS(fixed8, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/fixed8CAT.Rdata')

# now we check whether they converged
final_models8 <- list()
for (mdli in 1:length(fixed8)) {
  
  mdl <- fixed8[[mdli]]
  
  if ((isSingular(mdl) == FALSE)&(mdl@optinfo$conv$opt == 0L)&(is.null(mdl@optinfo$conv$lme4$code))) {
    final_models8 <- append(final_models8, mdl)
  }
}

# they did, so we pick the model with the highest AIC
fixed_aics = sapply(final_models8, AIC)
best_model8 = final_models8[order(fixed_aics, decreasing=FALSE)][[1]]
print(formula(best_model8))
summary(best_model8)
saveRDS(best_model8, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model8CAT.Rdata')
```

#### Permutation test
```{r word index 8 permutation, include=TRUE, message=FALSE, warning=FALSE}
best_model8 <- readRDS('C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/best_model8CAT.Rdata')
perm8 <- permute_lmer(best_model8,
                      subset(data, word_index==8),
                      n_permutations,
                      verbose=TRUE)
saveRDS(perm8, file='C:/Users/slaats/Documents/MPI/Project/Project 3 - Behavioural studies/3 - SPR main/analysis/perm8')

```

```{r word index 8 viz permutation, fig.width=7.5,fig.height=4, include=TRUE, message=FALSE, warning=FALSE}
plots8 <- plot_permutation_densities(perm8)
print(plots8)
print(perm8$p_values)
```
